Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to US Midwest agroecosystems

### 2021-08-26 Frontiers/Frontiers.tex
as

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome weed species in the United States. Palmer amaranth is endemic to Southern United States but its range is expanding northward. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted to investigate morphology, flowering and sex dimorphism from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations of United States Midwest. In general, results demonstrated that first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from second cohort (established in July). At first cohort, Palmer amaranth produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas Palmer amaranth produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} at second cohort in bareground, soybean and corn, respectively. The plant height was more impacted by Palmer amaranth growing in corn. Palmer amaranth plants averaged 85.2 cm tall in the first cohort but 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Also, Palmer amaranth sex dimorphism was slightly influenced by day of year, weight and height. The model estimated that probability of being a female plant increase as biomass and height increases. Our results showed the fast adaptation and plasticity of Palmer amaranth to grow and adapt to cropping systems from United States Midwest. Palmer amaranth is likely to continue its expansion northward. Therefore, preventing plant dispersal into new habitats is the most effective management strategy. Reactive management to reduce Palmer amaranth impact on cropping systems should encompass diversity of tactics that minimize the species ability establish into cropping systems, including crop rotation (not only corn and soybean), early/late crop planting, row spacing, and cover crops. Our results also shed a light on sex dimorphism as a future Palmer amaranth ecological management.

@@

Palmer amaranth is a fast growing summer annual forb indigenous to the
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with plant
biology. Farm mechanization, adoption of conservation agriculture (e.g.,
no-till), and intensive use of herbicides for weed management are the
main human-mediated selection of Palmer amaranth into cropping systems
(Ward et al., 2013). On the other hand, Palmer amaranth is a prolific
seed producer with a C4 photosynthetic apparatus (Wang et al., 1992).
With a dioecy nature, Palmer amaranth male and female plants are
obligate outcrosser species, increasing the chances of exchanging
adaptive traits among plants (Jhala et al., 2021; Oliveira et al.,
2018). Also, Palmer amaranth small seed size (e.g, 1 mm) tend to thrive
in no-tillage systems (Price et al., 2011), and spread across locations
through farm equipment (Sauer, 1972), manure (Hartzler and Anderson,
2016), animals (Farmer et al., 2017). The dispersal capacity of Palmer
amaranth makes the species one of the most successful cases of weed
adaptation to current cropping systems.

Light and temperature are the main environment requirements for Palmer
amaranth successful grow and development (Jha et al., 2010). Palmer
amaranth is reported with an extended germination period (Ward et al.,
2013). Germination of Palmer amaranth was triggered by 18 C soil
temperature at 5 cm depth (Keeley et al., 1987), and optimal germination
and biomass production occurred at 35/30 C day and night temperatures
(Guo and Al-Khatib, 2003). In addition, Palmer amaranth establishment is
human-mediated by tillage timings and preemergence-applied herbicides
(Chahal et al., 2021), which can result in weed germination shifts
(Sbatella and Wilson, 2010). Water has not shown to limit Palmer
amaranth fitness. Under continuous water stress, Palmer amaranth
survived and produced at least 14000 seeds plant\textsuperscript{-1}
(Chahal et al., 2018). Seeds from Palmer amaranth growing with limited
water conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). Growing conditions and management practices
also influence Palmer amaranth sex dimorphism and flowering pattern
(Korres et al., 2017; Rumpa et al., 2019). Therefore, Palmer amaranth
has shown plasticity to evolve and fast adapt under the current
conditions. Further scenarios show that global temperature warming can
impact agriculture, and promote niches for Palmer amaranth
invasion/adaptation into new environments. Currently, it is estimated
that the greatest climatic risk of Palmer amaranth establishment are
agronomic crops in Australia and Sub-Sahara Africa (Kistner and
Hatfield, 2018). Temperature is a key factor limiting Palmer amaranth
expansion to cooler geographies (Briscoe Runquist et al., 2019);
however, under future climate change Palmer amaranth is likely to expand
northward into Canada and Northern Europe (Kistner and Hatfield, 2018;
Briscoe Runquist et al., 2019).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the United States, Palmer amaranth is well established
in the Cotton Belt (Garetson et al., 2019; Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021). There are some reported
cases of Palmer amaranth in Michigan (Kohrt et al., 2017) and
Connecticut (Aulakh et al., 2021). Also, it is estimated that Palmer
amaranth can cause high damage to soybean fields in Illinois (Davis et
al., 2015), which is concerning as soybean along with corn make most of
US Midwest agronomic hectares. In Iowa, a study showed that Palmer
amaranth is still not well adapted compared to waterhemp
(\emph{Amaranthus tuberculatus}) (Baker, 2021). Waterhemp, an
\emph{Amaranthus} species, is the most troublesome species in that
geography (Tranel et al., 2011). Nonetheless, invasion and successful
eradication of Palmer amaranth is documented in Minnesota (Yu et al.,
2021). Palmer amaranth infestations have not been detected in Canada;
however, Palmer amaranth seeds were detected in sweet potato slips in
the country (Page et al., 2021). Palmer amaranth is still not as well
adapted and established to Northern as it is in the Southern North
America. Therefore, Palmer amaranth range of expansion into new habitats
can increase. It seems certain the need to manage Palmer amaranth in
agronomic crops throughout novel environments in the near future.
Strategies on Palmer amaranth management should encompass the
agroecosystem level but not only attempts to eradicate the weed. Most
tactics to manage Palmer amaranth are based on technology fixes (Scott,
2011), which are short-term (e.g., herbicide and/or tillage) rather than
long-term weed management. Palmer amaranth management should be built on
minimizing the species ability to adapt, grow and develop into
agroecossystems.

@@

A Palmer amaranth accession (Kei3) from Perkins County, Nebraska. with
no reported herbicide resistance according to Oliveira et al. (2021) was
selected for this study. Three weeks prior to the onset of the field
experiments, seeds were planted in plastic trays containing potting-mix.
Emerged seedlings (1 cm) were transplanted into 200
cm\textsuperscript{-3} plastic pots (a plant pot\textsuperscript{-1}).
Palmer amaranth seedlings were supplied with adequate water and kept
under greenhouse conditions at the University of Wisconsin-Madison,
University of Nebraska-Lincoln, and Western Illinois University; and
kept outdoors at the Perkins extension office in Grant, NE until the 2-3
leaf stage (5 to 8 cm height) when they were transported to the field.

@@

The experiment was conducted in 2018 and 2019 under field conditions at
five locations: Arlington, WI (43°18'N, 89°29'W), Clay Center, NE
(40.57'N, 9814'W), Grant, NE (XXX'N, XXX'W), Lincoln, NE (41.16'N,
96.42'W), and Macomb, IL (XXX'N, XXX'W).

Fields were conventionally tilled prior to crop planting. Corn hybrid
and soybean varieties were planted in 76-cm row spacing (Table 1).
Monthly mean air temperature and total precipitation were obtained using
Daymet weather data from June through September across the five
locations in 2018 and 2019 (Correndo et al., 2021) (Figure 1)

@@

 & Weed control & glyphosate\textsuperscript{a} / \emph{S}-metolachor\textsuperscript{b} & \emph{S}-metolachlor + trazine + mesotrione + bicyclopyrone\textsuperscript{c} &  &  & \\

 & Stage at 1\textsuperscript{st} cohort & V2-3 &  &  &  & \\

 & Stage at 2\textsuperscript{nd} cohort & V6-7 &  &  &  & \\

@@

\multirow{-7}{*}{\raggedright\arraybackslash } & Fertilization & N (46-0-0) at 157 kg ha\textsuperscript{-1} &  &  &  & \\

@@

 & Weed control & glyphosate / \emph{S}-metolachor & saflufenacil + imazethapyr\textsuperscript{d} + pyroxasulfone &  &  & \\

 & Stage at 1\textsuperscript{st} cohort & V1-2 &  &  &  & \\

 & Stage at 2\textsuperscript{nd} cohort & V5-6 &  &  &  & \\

@@

The field experimental units were three adjacent 9.1 m wide (12 rows at
76.2 cm row spacing) by 10.7 m long. The experimental design were
arranged in factorial design with three crops, two transplanting times
simulating two cohorts, repeated across five locations. Each
experimental unit was planted with corn, soybean, or kept under
bareground. The two transplanting timings were June 1 (first cohort) and
July 1 (second cohort). Palmer amaranth seedlings (potting mix + two
seedlings) were transplanted (6 cm deep and 8 cm wide). Forty-eight
plants were equidistantly placed (0.76 m apart) between rows within each
crop. After a week, one plant was eliminated and one was kept, resulting
in 24 plants per experimental unit and transplanting time (Figure 2).
When needed, Palmer amaranth plants were supplied with water during the
first week after transplanting.

After transplanting, Palmer amaranth flowering was monitored until the
end of the study. When a plant flowered, the day was recorded, plant sex
was identified (male or female), plant height was measured from soil
surface to the top of plant. Also, aboveground plant biomass was harvest
near soil surface and oven dried at 65 C until reaching constant weight
before weighing (g plant\textsuperscript{-1}) was recorded.

Plants had to be harvested at flowering because Palmer amaranth is
neither endemic in Wisconsin nor in Illinois. In our study, all
locations follow the methodology of plant harvest at flowering
initiation, except in Grant, NE. In this location, all Palmer amaranth
plants were harvest at once on July 06, 2018 and 2019 (first cohort),
and on August 17, 2018 and on July 31, 2019 (second cohort).

@@

The Palmer amaranth flowering timing was estimated as cumulative
flowering across all location. Dataset was nested but it did not include
Grant, NE. Palmer amaranth cumulative flowering estimation was
determined using an asymmetrical three parameter log logistic Weibull
model of the drc package (Ritz et al., 2015).

@@

Palmer amaranth sex dimorphism was fitted to a binary logistic
regression (Bangdiwala, 2018). Binary logistic regression is used for
predicting binary classes, such as the probability of a plant being
female in a dioecious species. Prior to the analysis, all missing values
were removed from the dataset. Also, data from Grant was not used in
this analysis due to the uniform plant harvesting at that location. The
complete dataset was splitted into 80\% train and 20\% test data. The
80\% train is used for the model training and the 20\% test is used for
checking model performance on unseen dataset. With 80\% dataset, a
binary response variable, male (0) and female (1), was fitted to a
generalized linear model (base R \emph{glm} function) including day of
year harvest, height, weight, crop and month as independent variables
(without interaction). The model family was binomial with a logit
function. The model fit was assessed through pseudo R-squared values
(McFadden, Cox and Snell, Cragg and Uhler) and likelihood ratio using
\emph{nagelkerke} function from ``rcompanion'' package (Mangiafico,
2021). The marginal effects computation was performed with Average
Marginal Effects (AMEs) at every observed value of x and average across
the results (Leeper, 2017) using \emph{margins} function from
``margins'' package (Leeper et al., 2021). The rest 20\% dataset was
predicted using \emph{predict} function with a cutoff estimation for
male or female using \emph{performance} function. The model quality
prediction from the classification algorithm was measured with precision
(\emph{precision} function), recall (\emph{recall} function) and
F1-score (\emph{f\_meas} function) using the ``yardstick'' package (Kuhn
et al., 2021). The precision determines the accuracy of positive
predictions (female plants), recall determines the fraction of positives
that were correctly identified, and F1-score is a weighted harmonic mean
of precision and recall with the best score of 1 and the worst score of
0. F1-score conveys the balance between the precision and the recall
(Yacouby and Axman, 2020).

@@

Palmer amaranth cumulative flowering at the second cohort ranged from
mid July to mid September (Figure 4B). Palmer amaranth growing in the
bareground resulted in earlier flowering time compared to soybean and
corn. Palmer amaranth growing in bareground reached 10\%, 50\%, and 90\%
flowering time at day 203.8, 214.4, and 232.2, respectively. Palmer
amaranth growing in soybean reached 10\% flowering at doy 210.9, which
was 6 days prior to corn (\emph{P}-value = 0.00). Similar trend was
observed at 50\% flowering, whereas Palmer amaranth reached 50\%
flowering in corn (doy 233.0) 4 days after soybeans (doy 228.9; \emph{P}
= 0.00). The 90\% Palmer amaranth cumulative flowering occurred at same
day in corn (260.9) and soybean (260.5; \emph{P} = 0.66).

@@

The model fit was 0.23, 0.32, 0.40 using pseudo R-squared test from
McFadden, Cox and Snell, and Cragg and Uhler, respectively. The
likelihood ratio test showed a p-value of \textless{} 0.00. The average
marginal effects showed that Palmer amaranth growing in corn resulted in
14.8\% less females plants (Table 2). Moreover, increasing a unit doy
increases the probability of having a female plant by 0.4\% (Table 2 and
Figure4A). Similar trend is observed for weight (Figure 4B) as well as
height (Figure 4C), whereas the probability of being female increase by
0.2\% and 0.1\% when a unit of weight (g) and height (cm) increases,
respectively.

@@

\caption{\label{tab:unnamed-chunk-3}Average marginal means of Palmer amaranth sex dimorphism logistic model. Factor pararemter values (crop and bareground) is shown compared to soybean.}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lrrrrrr}
\toprule
Term & AME\textsuperscript{a} & SE\textsuperscript{b} & Lower & Upper & Z-score & P-value\\
\midrule
crop\_bareground & -0.048 & 0.054 & -0.154 & 0.059 & -0.876 & 0.381\\

@@

The model accuracy evaluation accuracy in the 20\% test dataset was 0.62
with a cutoff value for female and male plants of 0.43. The model
classification showed a precision of 0.64, recall of 0.66, and a
F1-score of 0.65. In addition, the area under the curve was 0.64.

@@

Seed production was not evaluated due to plant harvest at initiation of
flowering. Nonetheless, it is well documented a strong positive
correlation between Palmer amaranth biomass and seed production
(Schwartz et al., 2016; Spaunhorst et al., 2018). I our study, plants
growing at first cohort accumulated an overall 42\% more biomass when
compared to second cohort. Therefore, Palmer amaranth plants growing in
the second cohort is likely to produce less seeds regardless the crop.
Our observation is consistent with the findings that first Palmer
amaranth cohort produced 50\% more seeds per plant than Palmer amaranth
plants established six weeks later in bareground (Webster and Grey,
2015). Still, seed production at second cohort is likely to replenish
the soil seedbank. Seed production and deposition in the seedbank is
also a key factor for species perpetuation (Menges, 1987). Palmer
amaranth can produce hundred thousands seeds per plant (Schwartz et al.,
2016; Keeley et al., 1987), and stay viable buried in the seedbank for
at least 36 months (Sosnoskie et al., 2013). Therefore, preventing
Palmer amaranth seed production or/and seed migration to its non-native
habitat is an essential strategy to minimize the species impact into
agroecossystem (Davis et al., 2015).

An ecological approach to reduce seed production in Palmer amaranth is
understanding its flowering window. Our study suggests that Palmer
amaranth flowering was slightly influenced by crops and cohort timings.
Palmer amaranth growing in bareground and corn resulted in the overall
shortest and longest flowering window, respectively. The shortest
flowering window of second cohorts of Palmer amaranth growing in
bareground highlighted the impact of photoperiod on flowering. When
growing in soybean, Palmer amaranth flowering window was similar to
bareground at first cohort but similar to corn at second cohort timing.
Plant flowering initiation is complex and depends on biological and
ecological factors (Lang, 1965). We hypothesize that when growing in
high competition (e.g., second cohort), Palmer amaranth plants tend to
initiate flowering early, as well as having an extended flowering
window. A study has shown that non-native Palmer amaranth initiated
flowering two weeks prior to the native waterhemp in Iowa (Baker, 2021).
Early flower initiation could be also a plant strategy when growing in
stress conditions. For example, when growing under water stress, early
flowering in Palmer amaranth resulted in a mismatch between female and
male plants by seven days (Mesgaran et al., 2021). A mismatch in Palmer
amaranth male and female flowering period can minimize plant outcross,
reducing plant seed production and/or exchange of resistant alleles
(Jhala et al., 2021). Sex dimorphism manipulation is considered a
potential ecological pest control (McFarlane et al., 2018; Schliekelman
et al., 2005).

>>>>

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome weed species in the United States. Palmer amaranth is endemic to Southern United States but its range is expanding northward. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted to investigate morphology, flowering and sex dimorphism from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations of US Midwest. In general, results demonstrated that first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from second cohort (established in July). The first Palmer amaranth first cohort produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas the second Palmer amaranth cohort produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} in bareground, soybean and corn, respectively. Palmer amaranth height was more impacted when growing in corn. Palmer amaranth plants averaged 85.2 cm tall in the first cohort but 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Also, Palmer amaranth sex dimorphism was slightly influenced by day of year, weight and height. The model estimated that probability of being a female plant increased as biomass and height increased. Our results showed the fast adaptation and plasticity of Palmer amaranth to grow and adapt to cropping systems from the US Midwest. Palmer amaranth is likely to continue its expansion northward. Therefore, preventing plant dispersal into new habitats is the most effective management strategy. Reactive management to reduce Palmer amaranth impact on cropping systems should encompass diversity of tactics that minimize the species ability to establish into cropping systems, including crop rotation (beyond corn and soybean), early/late crop planting, row spacing, cover crops, and effective chemical control programs.

@@

Palmer amaranth is a fast growing summer annual forb indigenous to the
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with plant
biology. Farm mechanization, adoption of conservation agriculture (e.g.,
no-till), and intensive use of herbicides for weed management are the
main human-mediated selection of Palmer amaranth into cropping systems
(Ward et al., 2013). On the other hand, Palmer amaranth is a prolific
seed producer with a C4 photosynthetic apparatus (Wang et al., 1992).
With a dioecy nature, Palmer amaranth male and female plants are
obligate outcrosser species, increasing the chances of exchanging
adaptive traits among plants (Jhala et al., 2021; Oliveira et al.,
2018). Also, Palmer amaranth small seeds (e.g, 1 mm) tend to thrive in
no-tillage systems (Price et al., 2011), and spread across locations
through farm equipment (Sauer, 1972), manure (Hartzler and Anderson,
2016), wildlife (Farmer et al., 2017), etc. The dispersal capacity of
Palmer amaranth makes the species one of the most successful cases of
weed adaptation to current cropping systems.

Light and temperature are the main environmental requirements for Palmer
amaranth growth and development (Jha et al., 2010). Palmer amaranth is
reported with an extended germination period (Ward et al., 2013).
Germination of Palmer amaranth was triggered by 18 C soil temperature at
5 cm depth (Keeley et al., 1987), and optimal germination and biomass
production occurred at 35/30 C day and night temperatures (Guo and
Al-Khatib, 2003). In addition, Palmer amaranth establishment is
human-mediated by tillage timings and preemergence-applied herbicides
(Chahal et al., 2021), which can result in weed germination shifts
(Sbatella and Wilson, 2010). Soil water content has not shown to limit
Palmer amaranth fitness. Under continuous water stress, Palmer amaranth
survived and produced at least 14000 seeds plant\textsuperscript{-1}
(Chahal et al., 2018). Seeds from Palmer amaranth growing with limited
water conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). Growing conditions and management practices
also influence Palmer amaranth sex dimorphism and flowering pattern
(Korres et al., 2017; Rumpa et al., 2019). Therefore, Palmer amaranth
has shown plasticity to evolve and fast adapt under the current
agroecosystem conditions. Further scenarios show that global temperature
warming can impact agriculture, and promote niches for Palmer amaranth
invasion/adaptation into new environments. Currently, it is estimated
that the greatest climatic risk of Palmer amaranth establishment are
agronomic crops in Australia and Sub-Sahara Africa (Kistner and
Hatfield, 2018). Temperature is a key factor limiting Palmer amaranth
expansion to cooler geographies (Briscoe Runquist et al., 2019);
however, under future climate change Palmer amaranth is likely to expand
northward into Canada and Northern Europe (Kistner and Hatfield, 2018;
Briscoe Runquist et al., 2019).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the United States, Palmer amaranth is well established
in the Cotton Belt (Garetson et al., 2019; Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021). There are some reported
cases of Palmer amaranth in Michigan (Kohrt et al., 2017) and
Connecticut (Aulakh et al., 2021). Also, it is estimated that Palmer
amaranth can cause high damage to soybean fields in Illinois (Davis et
al., 2015), which is concerning as soybean along with corn make most of
US Midwest agronomic hectares. In Iowa, a study showed that Palmer
amaranth is still not well adapted compared to waterhemp
(\emph{Amaranthus tuberculatus}) (Baker, 2021). Waterhemp, an
\emph{Amaranthus} species, is the most troublesome species in the US
Midwest (Tranel et al., 2011). Nonetheless, invasion and successful
eradication of Palmer amaranth is documented in Minnesota (Yu et al.,
2021). Palmer amaranth infestations have not been detected in Canada;
however, Palmer amaranth seeds were detected in sweet potato slips in
the country (Page et al., 2021). Palmer amaranth is still not as well
adapted and established to Northern as it is in the Southern North
America. Therefore, Palmer amaranth range of expansion into new habitats
can increase. It seems certain the need to manage new Palmer amaranth
infestations in agronomic crops throughout northern United States in the
near future. Strategies on Palmer amaranth management should encompass
the agroecosystem level but not only attempts to eradicate the weed.
Most tactics to manage Palmer amaranth are based on technology fixes
(Scott, 2011), which are short-term (e.g., herbicide and/or tillage)
rather than long-term weed management. Palmer amaranth management should
be built on minimizing the species ability to adapt, grow and develop
into agroecossystems.

@@

A Palmer amaranth accession (Per1) from Perkins County, Nebraska with no
reported herbicide resistance according to Oliveira et al. (2021) was
selected for this study. Three weeks prior to the establishment of each
cohort, seeds were planted in plastic trays containing potting-mix.
Emerged seedlings (1 cm) were transplanted into 200
cm\textsuperscript{-3} plastic pots (a plant pot\textsuperscript{-1}).
Palmer amaranth seedlings were supplied with adequate water and kept
under greenhouse conditions at the University of Wisconsin-Madison,
University of Nebraska-Lincoln, and Western Illinois University; and
kept outdoors at the Perkins extension office in Grant, NE until the 2-3
leaf stage (5 to 8 cm height) when they were transported to the field.

@@

The experiment was conducted in 2018 and 2019 under field conditions at
five locations: Arlington, WI (43°18'N, 89°29'W), Clay Center, NE
(40.57'N, 9814'W), Grant, NE ('N, 'W), Lincoln, NE (41.16'N, 96.42'W),
and Macomb, IL ('N, 'W).

Fields were conventionally tilled prior to crop planting. Corn and
soybean were planted in 76-cm row spacing (Table 1). Monthly mean air
temperature and total precipitation were obtained using Daymet weather
data from June through September across the five locations in 2018 and
2019 (Correndo et al., 2021) (Figure 1)

@@

 & Weed control & glyphosate / S-metolachor & S-metolachlor + trazine + mesotrione, + bicyclopyrone &  &  & \\

 & Stage at 1 cohort & V2-3 &  &  &  & \\

 & Stage at 2 cohort & V6-7 &  &  &  & \\

@@

\multirow{-7}{*}{\raggedright\arraybackslash } & Fertilization & N (46-0-0) at 157 kg ha-1 &  &  &  & \\

@@

 & Weed control & glyphosate / S-metolachor & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

 & Stage at 1 cohort & V1-2 &  &  &  & \\

 & Stage at 2 cohort & V5-6 &  &  &  & \\

@@

The field experimental units were three adjacent 9.1 m wide (12 rows at
76.2 cm row spacing) by 10.7 m long. The experimental design were
arranged in factorial design with three crops, two transplanting times
simulating two cohorts, repeated across five locations. Each filed
experimental unit was planted with corn, soybean, or kept under
bareground. The two transplanting timings were June 1 (first cohort) and
July 1 (second cohort). Palmer amaranth seedlings (potting mix + two
seedlings) were transplanted (6 cm deep and 8 cm wide). Forty-eight
plants were equidistantly placed (0.76 m apart) between rows within each
crop. After a week, one plant was eliminated and one was kept, resulting
in 24 plants per experimental unit and transplanting time (Figure 2).
When needed, Palmer amaranth plants were supplied with water during the
first week after transplanting to assure seedling survival.

After transplanting, Palmer amaranth flowering was monitored until the
end of the study. When a plant flowered, the day was recorded, plant sex
was identified (male or female), plant height was measured from soil
surface to the top of plant. Also, aboveground plant biomass was
harvested near soil surface and oven dried at 65 C until reaching
constant weight before weigh (g plant\textsuperscript{-1}) was recorded.

Plants had to be harvested at flowering because Palmer amaranth is
neither endemic in Wisconsin nor in Illinois. In our study, all
locations followed the methodology of plant harvest at flowering
initiation, except in Grant, NE. In this location, all Palmer amaranth
plants were harvest at once on July 06, 2018 and 2019 (first cohort),
and on August 17, 2018 and on July 31, 2019 (second cohort).

@@

The Palmer amaranth flowering timing was estimated as cumulative
flowering across all locations, except Grant, NE. Palmer amaranth
cumulative flowering estimation was determined using an asymmetrical
three parameter log logistic Weibull model of the drc package (Ritz et
al., 2015).

@@

A binary logistic regression was fitted to Palmer amaranth sex
dimorphism. Binary logistic regression is used for predicting binary
classes (Bangdiwala, 2018), such as the probability of a plant being
female in a dioecious species. Prior to the analysis, missing values
were removed from the dataset. Also, data from Grant was not used in
this analysis due to the uniform plant harvesting at that location. The
complete dataset was splitted into 80\% train and 20\% test data. The
80\% train is used for the model training and the 20\% test is used for
checking model performance on unseen dataset. With 80\% dataset, a
generalized linear model (base R \emph{glm} function) was fitted to
binary response variable, the probability of being female (0 to male and
1 to female). The independent variables were day of year harvest,
height, weight, and crop (without interaction). The model family was
binomial with a logit function. The model fit was assessed through
pseudo R-squared values (McFadden, Cox and Snell, Cragg and Uhler) and
likelihood ratio using \emph{nagelkerke} function from ``rcompanion''
package (Mangiafico, 2021). The marginal effects computation was
performed with Average Marginal Effects (AMEs) at every observed value
of x and average across the results (Leeper, 2017) using \emph{margins}
function from ``margins'' package (Leeper et al., 2021). The 20\% test
data was predicted using the \emph{predict} function with a cutoff
estimation for male or female using \emph{performance} function. The
model quality prediction from the classification algorithm was measured
with precision (\emph{precision} function), recall (\emph{recall}
function) and F1-score (\emph{f\_meas} function) using the ``yardstick''
package (Kuhn et al., 2021). The precision determines the accuracy of
positive predictions (female plants), recall determines the fraction of
positives that were correctly identified, and F1-score is a weighted
harmonic mean of precision and recall with the best score of 1 and the
worst score of 0. F1-score conveys the balance between the precision and
the recall (Yacouby and Axman, 2020). The area under the curve was also
estimated with XXXXXXXXXXXXXXXXX.

@@

Palmer amaranth cumulative flowering in the second cohort ranged from
mid July to mid September (Figure 4B). Palmer amaranth growing in the
bareground resulted in earlier flowering time compared to soybean and
corn. Palmer amaranth growing in bareground reached 10\%, 50\%, and 90\%
flowering time at day 203.8, 214.4, and 232.2, respectively. Palmer
amaranth growing in soybean reached 10\% flowering at doy 210.9, which
was 6 days prior to corn (\emph{P}-value = 0.00). Similar trend was
observed at 50\% flowering, whereas Palmer amaranth reached 50\%
flowering in corn (doy 233.0) 4 days after soybeans (doy 228.9; \emph{P}
= 0.00). The 90\% Palmer amaranth cumulative flowering occurred at same
day in corn (260.9) and soybean (260.5; \emph{P} = 0.66).

@@

The model goodness of fit was 0.23, 0.32, 0.40 using pseudo R-squared
test from McFadden, Cox and Snell, and Cragg and Uhler, respectively.
The likelihood ratio test showed a p-value of \textless{} 0.00. The
average marginal effects showed that Palmer amaranth growing in corn
resulted in 14.8\% less females plants (Table 2). Moreover, increasing a
unit doy increases the probability of having a female plant by 0.4\%
(Table 2 and Figure4A). Similar trend is observed for weight (Figure 4B)
as well as height (Figure 4C), whereas the probability of being female
increase by 0.2\% and 0.1\% when a unit of weight (g) and height (cm)
increases, respectively.

@@

\caption{\label{tab:unnamed-chunk-3}Average marginal effects of Palmer amaranth sex dimorphism logistic model. Factor pararemter values (crop and bareground) is shown compared to soybean.}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lrrrrrr}
\toprule
Term & AME & SE & Lower & Upper & Z-score & P-value\\
\midrule
crop\_bareground & -0.048 & 0.054 & -0.154 & 0.059 & -0.876 & 0.381\\

@@

The model accuracy evaluation accuracy in the 20\% test dataset was 0.62
with a cutoff value for female and male plants of 0.43. The model
classification showed a precision of 0.64, recall of 0.66, and a
F1-score of 0.65. In addition, the AUC was 0.64.

@@

Seed production was not evaluated due to plant harvest at initiation of
flowering. Nonetheless, it is well documented a strong positive
correlation between Palmer amaranth biomass and seed production
(Schwartz et al., 2016; Spaunhorst et al., 2018). I our study, plants
growing at first cohort accumulated an overall 42\% more biomass when
compared to second cohort. Therefore, Palmer amaranth plants growing in
the second cohort is likely to produce less seeds regardless the crop.
Our observation is consistent with the findings that first Palmer
amaranth cohort produced 50\% more seeds per plant than Palmer amaranth
plants established six weeks later in bareground (Webster and Grey,
2015). Still, seed production at second cohort is likely to replenish
the soil seedbank. Seed production and deposition in the seedbank is
also a key factor for species perpetuation (Menges, 1987). Palmer
amaranth can produce hundred thousands seeds per plant (Schwartz et al.,
2016; Keeley et al., 1987), which can stay viable in the soil seedbank
for at least 36 months (Sosnoskie et al., 2013). Therefore, preventing
Palmer amaranth seed production or/and seed migration to non-native
habitats is an essential strategy to minimize the species impact in
agroecosystems (Davis et al., 2015).

An ecological approach to reduce Palmer amaranth seed production is
understanding plant biology, including flowering pattern. Our study
suggests that Palmer amaranth flowering was slightly influenced by crops
and cohort timings. Palmer amaranth growing in bareground and corn
resulted in the overall shortest and longest flowering window,
respectively. The shortest flowering window of second cohorts of Palmer
amaranth growing in bareground highlighted the impact of photoperiod on
flowering. When growing in soybean, Palmer amaranth flowering window was
similar to bareground at first cohort but similar to corn at second
cohort timing. Plant flowering initiation is complex and depends on
biological and ecological factors (Lang, 1965). We hypothesize that when
growing in high competition (e.g., second cohort), Palmer amaranth
plants tend to initiate flowering early, as well as having an extended
flowering window. A study has shown that non-native Palmer amaranth
initiated flowering two weeks prior to the native waterhemp in Iowa
(Baker, 2021). Early flower initiation could also be a plant strategy
when growing in stress conditions. For example, when growing under water
stress, early flowering in Palmer amaranth resulted in a mismatch
between female and male plants by seven days (Mesgaran et al., 2021). A
mismatch in Palmer amaranth male and female flowering period can
minimize plant outcross, reducing plant seed production and/or exchange
of resistant alleles (Jhala et al., 2021). Sex dimorphism manipulation
is considered a potential ecological pest control (McFarlane et al.,
2018; Schliekelman et al., 2005).

The mechanisms of sex-determination in plant species is intriguing and
arouse the curiosity of many scientists, including Charles Darwin
(Darwin, 1888). In our study, the sex dimorphism model performance was
decent (AIC 0.64) considering the biology of plant flowering. A 1:1 male
and female sex ratio is a general evolutionary stable strategy for plant
species perpetuation (Fisher, 1930). However, a slight deviation from
1:1 sex ratio might occur in some dioecious species. For example, the
dioecious \emph{Halophila stipulacea} is a female-biased plant in its
native habitat, but the naturalized \emph{H. stipulacea} have a 1:1
ratio (Nguyen et al., 2018). Naturalized of \emph{H. stipulacea} reduced
female-male ratio to expand into its non-native habitat (Nguyen et al.,
2018). Also, biotic and/or abiotic stress can influence plant sex
determination. Palmer amaranth male-to-female ratio was greater under
high plant densities (Korres and Norsworthy, 2017) and after herbicide
application (Rumpa et al., 2019). Our model estimated that late
flowering, heavier and taller Palmer amaranth plants deviated from 1:1
ratio in favor to female plants. It was reported that female Palmer
amaranth plants invested more in height, stem and biomass while male
invested more in leaf area and leaf dry weight under nutrient deficiency
(Korres et al., 2017). Our model also estimated more female plants in
soybean and bareground compared to corn, which might be linked to plant
competition strategy in each crop. Our results showed the influence of
life-history and ecological traits on sexual dimorphism in Palmer
amaranth. Sexual dimorphism is documented in other dioecious species
(Barrett and Hough, 2013). For example, stronger female plant
competition and greater male tolerance to herbivory was reported in
\emph{Spinacia oleracea} (Pérez-Llorca and Sánchez Vilas, 2019).
Research on candidate genes for sex determination in \emph{Amaranthus}
species are currently underway but it is far from complete (Montgomery
et al., 2021, 2019). Further studies are also needed to understand the
ecological basis of Palmer amaranth flowering, including the plant
behavior under climate change.
